%\documentclass[letterpaper,11pt]{article}
\documentclass[UTF8]{ctexart}
\usepackage[hang,flushmargin]{footmisc}

%\usepackage[empty]{fullpage}
\pagestyle{plain}

\setlength{\topmargin}{-15pt}
\setlength{\oddsidemargin}{0pt}
\setlength{\evensidemargin}{0pt}
\setlength{\textwidth}{460pt}
\setlength{\voffset}{-1.5cm}
\setlength{\textheight}{23cm}
\setlength{\parindent}{0pt}
\setlength{\parskip}{10pt}

\nofiles

\usepackage{gb4e}

\setlength{\glossglue}{5pt plus 2pt minus 1pt}
\renewcommand{\eachwordthree}{\rule[-10pt]{0pt}{0pt}}

\begin{document}
\begin{center}
\textbf{\Large{Le Petit Prince - Glosses from Chinese}}\\
%\vspace{3pt}
%\large{Sean Zhang}\\
%xiaoronz@andrew.cmu.edu
\end{center}

\subsection*{Chapter 1}

\begin{exe}
\ex 
\glll
当 我 还 只 有 六 岁 的 时候 ， 在 一 本 描写 原始 森林 的 名 叫 《 真实 的 故事 》 的 书 中 ， 看 到 了 一 副 精彩 的 插画 ， 画 的 是 一 条 蟒蛇 正在 吞食 一 只 大 野兽 。
\\
dang1 wo3 hai2 zhi2 you3 liu4 sui4 de5 shi2hou5 , zai4 yi4 ben3 miao2xie3 yuan2shi3 sen1lin2 de5 ming2 jiao4 《 zhen1shi2 de5 gu4shi4 》 de5 shu1 zhong1 , kan4 dao4 le yi2 fu4 jing1cai3 de5 cha1hua4 , hua4 de5 shi4 yi4 tiao2 mang3she2 zheng4zai4 tun1shi2 yi4 zhi1 da4 ye3shou4 。
\\
When \textsc{1sg} yet only have 6 years-old DE time , at one \textsc{clf} depict primitive forest DE name called 《 real DE story 》 DE book in , see reach LE one \textsc{clf} magnificent DE illustration , painted DE is one \textsc{clf} boa \textsc{prog} swallow one \textsc{clf} big beast . 
\\
\trans When I was only 6 years old, in a book depicting primitive forests called ``real stories'', I saw a magnificent illustration showing a boa swallowing a giant beast. 
\ex
\glll
页头 上 就 是 那 副 画 的 摹本 。 
\\
ye4tou2 shang4 jiu4 shi4 na4 fu2 hua4 de5 mo2ben3 。
\\
{page header} on exactly is that \textsc{clf} painting DE copy .
\\
\trans A copy of that painting is at the header of this page. 
\ex
\glll
这 本 书 中 写 道 ： `` 这些 蟒蛇 把 它们 的 猎获物 不 加 咀嚼 地 囫囵 吞下 ， 尔后 就 不 能 再 动弹 了 ； 它们 就 在 长长 的 六 个 月 的 睡眠 中 消化 这些 食物 。 '' 
\\
zhe4 ben3 shu1 zhong1 xie3 dao4 : `` zhe4xie1 mang3she2 ba2 ta1men5 de5 lie4huo4wu4 bu4 jia1 ju3jue2 de5 hu2lun2 tun1xia4 , er3hou4 jiu4 bu4 neng2 zai4 dong4tan2 le5 ; ta1men5 jiu4 zai4 chang2chang2 de5 liu4 ge4 yue4 de5 shui4mian2 zhong1 xiao1hua4 zhe1xie1 shi2wu4 。 ''
\\
this \textsc{clf} book inside write \textsc{comp} : `` these boa BA they DE prey no add chew DE complete swallow , afterwards then not {able to} {continue to} move LE ; they just at long DE six \textsc{clf} month DE sleep inside digest these food . ''
\\
\trans In this book it says: ``these boas swallow their preys as a whole, without chewing, afterwards they are no longer able to move; they then digest these foods during a long hibernation of six months. 
\ex
\glll
当时 ， 我 对 丛林 中 的 奇遇 想 得 很 多 ， 于是 ， 我 也 用 彩色 铅笔 画 出 了 我 的 第一 副 图画 。 
\\
dang1shi2 , wo3 dui4 cong2lin2 zhong1 de5 qi2yu4 xiang3 de5 hen3 duo1 , yu2shi4 , wo2 ye3 yong4 cai3se4 qian1bi3 hua4 chu1 le5 wo3 de5 di4yi1 fu4 tu2hua4 。
\\
{at that time} , \textsc{1sg} regarding jungle inside DE adventure think DE very much , hence , \textsc{1sg} also use color pencil draw out LE \textsc{1sg} DE first \textsc{clf} painting .
\\
\trans At that time, I thought a lot about the adventures inside jungles, so I also used color pencils and drew my first painting. 
\ex
\glll
我 的 第一 号 作品 。 
\\
wo3 de5 di4yi1 hao4 zuo4pin3 。 
\\
\textsc{1sg} DE first \textsc{clf} work .
\\
\trans My first work. 
\ex
\glll
它 是 这样 的 ：
\\
ta1 shi4 zhe4yang4 de5 : 
\\
\textsc{3sg.inanim} is {like this} DE :
\\
\trans It is like this:
\ex
\glll
我 把 我 的 这 副 杰作 拿 给 大人 看 ， 我 问 他们 我 的 画 是不是 叫 他们 害怕 。
\\
wo2 ba3 wo3 de5 zhe4 fu2 jie2zuo4 na2 gei3 da4ren2 kan4 , wo3 wen4 ta1men5 wo3 de5 hua4 shi4bu5shi4 jiao4 ta1men5 hai4pa4 。
\\
\textsc{1sg} BA \textsc{1sg} DE this \textsc{clf} masterpiece take to adult see , \textsc{1sg} ask them \textsc{1sg} DE painting {is or not} make them afraid .
\\
\trans I showed this masterpiece to the adults, and I asked them whether or not my painting makes them afraid.
\ex
\glll
他们 回答 我 说 ： “ 一 顶 帽子 有 什么 可怕 的 ？ ”
\\
ta1men5 hui2da2 wo3 shuo1 : `` yi4 ding3 mao4zi5 you2 shen3me5 ke3ba4 de5 ? ''
\\
\textsc{3pl} answer me saying : `` one \textsc{clf} hat has what scary DE ? ''
\\
\trans They answered me saying: ``what's so scary about a hat?''
\ex
\glll
我 画 的 不 是 帽子 ， 是 一 条 巨蟒 在 消化 着 一 头 大象 。
\\
wo3 hua4 de5 bu2 shi4 mao4zi5 , shi4 yi4 tiao2 ju4mang2 zai4 xiao1hua4 zhe5 yi4 tou2 da4xiang4 。
\\
\textsc{1sg} draw DE not is hat , is one \textsc{clf} {giant boa} ZAI digest ZHE one \textsc{clf} elephant . 
\\
\trans What I drew is not a hat, it's a giant boa digesting an elephant. 
\ex
\glll
于是 我 又 把 巨蟒 肚子 里 的 情况 画 了 出来 ， 以 便 让 大人 们 能够 看懂 。
\\
yu2shi4 wo3 you4 ba3 ju4mang2 du4zi5 li3 de5 qing2kuang4 hua4 le5 chu1lai2 , yi3 bian4 rang4 da4ren2 men5 neng2gou4 kan4dong3 . 
\\
so \textsc{1sg} also BA boa stomach inside DE situation draw LE out , {} {so that} let adult \textsc{pl} {are able to} {understand} . 
\\
\trans So I also drew out the situation inside the giant boa's stomach, so that the adults can understand. 
\ex
\glll
这些 大人 总是 需要 解释 。
\\
zhe4xie1 da4ren2 zong3shi4 xu1yao4 jie3shi4 . 
\\
these adults always need explanations .
\\
\trans These adults always need explanations.
\ex
\glll
我 的 第二 号 作品 是 这样 的 ：
\\
wo3 de5 di4er4 hao4 zuo4pin2 shi4 zhe4yang4 de5 :
\\
\textsc{1sg} DE second \textsc{clf} work is {like this} DE:
\\
\trans My second work is like this:
\ex
\glll
大人 们 劝 我 把 这些 画 着 开 着 肚皮 的 ， 或 闭 上 肚皮 的 蟒蛇 的 图画 放 在 一边 ， 还是 把 兴趣 放 在 地理 、 历史 、 算术 、 语法 上 。
\\
da4ren2 men5 quan4 wo3 ba3 zhe4xie1 hua4 zhe5 kai1 zhe5 du4pi2 de5 ， huo4 bi4 shang4 du4pi2 de5 mang2she2 de5 tu2hua4 fang4 zai4 yi4bian1 ， hai2shi4 ba3 xing4qu4 fang4 zai4 di4li3 、 li4shi3 、 suan4shu4 、 yu2fa3 shang4 。
\\
adult \textsc{pl} advise \textsc{1sg} BA these draw ZHE open ZHE belly DE , or close up belly DE boa DE painting put at side , rather BA interest put at geography , history , arithmetic , grammar on .
\\
\trans The adults advised me to set aside the paintings of boas with belly open or with belly closed, and instead focus on geography, history, arithmetic, and grammar. 
\ex
\glll
就 这样 ， 在 六 岁 的 那 年 ， 我 就 放弃 了 当 画家 这 一 美好 的 职业 。
\\
jiu4 zhe4yang4 ， zai4 liu4 sui4 de5 na4 nian2 ， wo3 jiu4 fang4qi4 le5 dang1 hua4jia1 zhe4 yi1 mei2hao3 de4 zhi2ye4 。
\\
just {like this} , at six {years old} DE that year , \textsc{1sg} just abandon LE becoming painter this one wonderful DE career .
\\
\trans Just like that, at the age of six, I gave up the wonderful career of becoming a painter.
\ex
\glll
我 的 第一 号 、 第二 号 作品 的 不成功 ， 使 我 泄 了 气 。
\\
wo3 de5 di4yi1 hao4 、 di4er4 hao4 zuo4pin2 de5 bu4cheng2gong1 ， shi2 wo3 xie4 le5 qi4 。
\\
\textsc{1sg} DE first \textsc{clf} , second \textsc{pl} work DE failure , make \textsc{1sg} disheartened LE {} .
\\
\trans The failure of my first and second works made me disheartened.
\ex
\glll
这些 大人 们 ， 靠 他们 自己 什么 也 弄 不 懂 ， 还得 老是 不断 地 给 他们 作 解释 。
\\
zhe4xie1 da4ren2 men5 ， kao4 ta1men5 zi4ji3 shen3me5 ye3 nong4 bu4 dong3 ， hai2shi4 lao3shi4 bu2duan4 de5 gei3 ta1men5 zuo4 jie3shi4 。
\\
these adult \textsc{pl} , {depend on} they self what also {} not understand , still always {non-stop} DE give them make explanation .
\\
\trans These adults, they understand nothing on their own, and I always explain things for them. 
\ex
\glll
这 真 叫 孩子 们 腻味 。
\\
zhe4 zhen1 jiao4 hai2zi5 men5 ni4wei4 。
\\
this really make kid \textsc{pl} {fed up}.
\\
\trans This really makes the kids feel fed up. 
\ex
\glll
后来 ， 我 只好 选择 了 另外 一 个 职业 ， 我 学会 了 开 飞机 ， 世界 各地 差不多 都 飞 到 过 。
\\
hou4lai2 ， wo3 zhi2hao3 xuan3ze2 le5 ling4wai4 yi2 ge4 zhi2ye4 ， wo3 xue2hui4 le5 kai1 fei1ji1 ， shi4jie4 ge4di4 cha1bu4duo1 dou1 fei1 dao4 guo4 。
\\
afterwards , \textsc{1sg} {can only} choose LE another one \textsc{clf} career , \textsc{1sg} learn LE drive plane , world everywhere almost all fly reach GUO .
\\
\trans Afterwards, I had to choose another career. I learned to fly a plane, and I've been to almost every place in the world. 
\ex
\glll
的确 ， 地理学 帮 了 我 很 大 的 忙 。
\\
di2que4 ， di4li3xue2 bang1 le5 wo2 hen3 da4 de5 mang2 。
\\
indeed , geography help LE \textsc{1sg} very big DE favor .
\\
\trans Indeed, geography helped me a lot. 
\ex
\glll
我 一眼 就 能 分辨 出 中国 和 亚里桑那 。
\\
wo3 yi4yan3 jiu4 neng2 fen1bian4 chu1 zhong1guo2 he2 ya4li3sang1na4 。
\\
\textsc{1sg} {in one glance} just can recognize {} China and Arizona .
\\
\trans I can recognize China and Arizona in one glance.
\ex
\glll
要是 夜里 迷失 了 航向 ， 这 是 很 有用 的 。
\\
yao4shi4 ye4li3 mi2shi1 le5 hang2xiang4 ， zhe4 shi4 hen2 you3yong4 de5 。
\\
if {at night} lose LE direction , this is very useful DE .
\\
\trans If I lost the direction at night, it is very useful. 
\ex
\glll
这样 ， 在 我 的 生活 中 ， 我 跟 许多 严肃 的 人 有 过 很 多 的 接触 。
\\
zhe4yang4 ， zai4 wo3 de5 sheng1huo2 zhong1 ， wo3 gen1 xu3duo1 yan2su4 de5 ren2 you3 guo4 hen3 duo1 de5 jie1chu4 。
\\
{like this} , in \textsc{1sg} DE life inside , \textsc{1sg} with many serious DE person have GUO very many DE contact .
\\
\trans Like that, in my life, I have had a lot of contact with many serious people. 
\ex
\glll
我 在 大人 们 中间 生活 过 很 长 时间 。
\\
wo3 zai4 da4ren2 men5 zhong1jian1 sheng1huo2 guo4 hen3 chang2 shi2jian1 。
\\
\textsc{1sg} in adult \textsc{pl} middle live GUO very long time .
\\
\trans I have lived for a long time among the adults. 
\ex
\glll
我 仔细 地 观察 过 他们 ， 但 这 并 没有 使 我 对 他们 的 看法 有 多大 的 改变 。
\\
wo2 zi3xi4 de5 guan1cha2 guo4 ta1men5 , dan4 zhe4 bing4 mei2you3 shi2 wo3 dui4 ta1men5 de5 kan4fa3 you3 duo1da4 de5 gai3bian4 。
\\
\textsc{1sg} carefully DE observe GUO \textsc{3pl} , but this {in fact} not make \textsc{1sg} regarding they DE opinion have big DE change .
\\
\trans I observed them carefully, but that in fact didn't change my opinion of them by much. 
\ex
\glll
当 我 遇到 一 个 头脑 看来 稍微 清楚 的 大人 时 ， 我 就 拿出 一直 保存 着 的 我 那 第一 号 作品 来 测试测试 他 。
\\
dang1 wo3 yu4dao4 yi2 ge4 tou2nao3 kan4lai2 bi3jiao4 qing1chu3 de5 da4ren2 shi2 ， wo3 jiu4 na2chu1 yi1zhi2 bao3cun2 zhe5 de5 wo3 na4 di4yi1 hao4 zuo4pin3 lai2 ce4shi4ce4shi4 ta1 。
\\
when \textsc{1sg} encounter one CLF mind seem rather  clear  DE adult time , \textsc{1sg} just {take out} always keep   ZHE DE \textsc{1sg}  that first CLF  work  to   test-test \textsc{3sg.m} .
\\
\trans When I encounter an adult whose mind seems to be rather clear, I take out my work number one that I've been always keeping to test him.
\ex
\glll
我 想 知道 他 是否 真的 有 理解 能力 。
\\
wo2 xiang3 zhi1dao4 ta1 shi4fou3 zhen1de5 you3 li2jie3 neng2li4 。
\\
I {want to} know \textsc{3sg.m} {yes or no} really have comprehension ability .
\\
\trans I wanted to know if he really had the ability to comprehend. 
\ex
\glll
可是 ， 得到 的 回答 总 是 ： “ 这 是 顶 帽子 。 ”
\\
ke3shi4 ， de2dao4 de5 hui3da2 zong3 shi4 ： “ zhe4 shi4 ding3 mao4zi5 。 ”
\\
however , receive DE response always is : `` this is \textsc{clf} hat . ''
\\
\trans But the response I got was always: ``this is a hat.''
\ex
\glll
我 就 不 和 他 谈 巨蟒 呀 ， 原始 森林 呀 ， 或者 星星 之类 的 事 。
\\
wo3 jiu4 bu4 he2 ta1 tan2 ju4mang3 a1 ， yuan2shi3 sen1lin2 a1 ， huo4zhe3 xing1xing1 zhi1lei4 de5 shi4 。
\\
\textsc{1sg} just not with \textsc{3sg.m} {talk about} boa \textsc{part} , primitive forest \textsc{part} , or stars {and the like} DE thing .
\\
\trans So I don't talk to him about things like boa, primitive forest, or stars. 
\ex
\glll
我 只得 迁就 他 的 水平 ， 和 他们 谈 些 桥牌 呀 ， 高尔夫球 呀 ， 政治 呀 ， 领带 呀 这些 。
\\
wo2 zhe3de2 qian1jiu4 ta1 d5e shui3ping2 ， he2 ta1men5 tan2 xie1 qiao2pai2 a1 ， gao1er3fu1qiu2 a1 ， zheng4zhi4 a1 ， ling3dai4 a1 zhe4xie1 。
\\
\textsc{1sg} {have to} {accommodate myself to} \textsc{3sg.m} DE level , with them  talk some bridge \textsc{part} , golf \textsc{part} , politics \textsc{part} , necktie \textsc{part} these .
\\
\trans I had to accommodate myself to his level, and talk to them about things like bridge, golf, politics, or necktie. 
\ex
\glll
于是 大人 们 就 十分 高兴 能 认识 我 这样 一 个 通情达理 的 人 。
\\
yu2shi4 da4ren2 men5 jiu4 shi2fen1 gao1xing4 neng2 ren4shi2 wo3 zhe4yang4 yi2 ge4 tong1qing2da2li3 de5 ren2 。
\\
so adult \textsc{pl} just very happy can {get to know} \textsc{1sg}  {like this} one \textsc{clf} reasonable DE person
\\
\trans So the adults are very have to get to know a reasonable person like me. 
\end{exe}
\newpage 

\subsection*{Chapter 2}
\begin{exe}
\ex
\glll
我 就 这样 孤独 地 生活 着 ， 没有 一 个 能 真正 谈 得 来 的 人 ， 一直 到 六 年 前 在 撒哈拉 沙漠 上 发生 了 那 次 故障 。
\\
wo3 jiu4 zhe4yang4 gu1du1 de5 sheng1huo2 zhe5 ， mei2you3 yi2 ge4 neng2 zhen1zheng4 tan2 de5 lai2 de5 ren2 ， yi1zhi2 dao4 liu4 nian2 qian2 zai4 sa4ha1la1 sha1muo4 shang4 fa1sheng1 le5 na4 ci4 gu4zhang4 。
\\
\textsc{1sg} just {like this} alone DE live ZHE , {not have} one \textsc{clf} can really talk DE {well} DE person , until at six years ago at Sahara desert on happen LE that \textsc{clf} accident . 
\\
\trans Just like this, I lived alone, without a person with whom I can get along well, until six years ago that accident happened at the Sahara desert. 
\ex
\glll
我 的 发动机 里 有 个 东西 损坏 了 。
\\
wo3 de5 fa1dong4ji1 li3 you3 ge4 dong1xi5 sun3huai4 le5 。
\\
\textsc{1sg} DE motor inside have \textsc{clf} something damaged LE .
\\
\trans Something was damaged inside my motor. 
\ex
\glll
当时 由于 我 既 没有 带 机械师 也 没有 带 旅客 ， 我 就 试图 独自 完成 这 个 困难 的 维修 工作 。
\\
dang1shi2 you2yu2 wo3 ji4 mei2you3 dai4 ji1xie4shi1 ye3 mei2you3 dai4 lv3ke4 ， wo3 jiu4 shi4tu2 du2zi4 wan2cheng2 zhe4 ge4 kun4nan2 de5 wei2xiu1 gong1zuo4 。
\\
{at that time} since \textsc{1sg} neither not bring mechanic nor not bring passenger , \textsc{1sg} just try alone finish this \textsc{clf} difficult DE repair operation .
\\
\trans Since at that time I had neither a mechanic nor a passenger with me, I tried to finish this difficult repair operation by myself. 
\ex
\glll
这 对 我 来说 是 个 生 与 死 的 问题 。
\\
zhe4 dui4 wo3 lai2shuo1 shi4 ge4 sheng1 yu2 si3 de5 wen4ti2 。
\\
this for \textsc{1sg} {} is \textsc{clf} life and death DE question .
\\
\trans To me this is a question of life and death. 
\ex
\glll
我 随身 带 的 水 只 够 饮用 一 星期 。
\\
wo3 sui2shen1 dai4 de5 shui3 zhi3 gou4 yin3yong4 yi4 xing1qi1 。
\\
\textsc{1sg} {with me} carry DE water only {be enough} drink one week .
\\
\trans The water that I carried with me is only enough to drink for one week. 
\ex
\glll
第一 天 晚上 我 就 睡 在 这 远离 人间 烟火 的 大 沙漠 上 。
\\
di4yi1 tian1 wan2shang4 wo3 jiu4 shui4 zai4 zhe4 yuan3li2 ren2jian1 yan1huo3 de5 da4 sha1muo4 shang4 。
\\
first day night \textsc{1sg} just sleep at this {far away from} human fire DE big desert above .
\\
\trans The first night, I slept at this big desert far away from human habitation.
\ex
\glll
我 比 大海 中 伏 在 小 木排 上 的 遇难者 还 要 孤独 得 多 。
\\
wo2 bi3 da4hai3 zhong1 fu2 zai4 xiao3 mu4pai2 shang4 de5 yu4nan4zhe3 hai2 yao4 gu1du1 de5 duo1 。
\\
\textsc{1sg} {compared to} ocean inside lie at small raft above DE castaway even am lonely DE much .
\\
\trans I'm much more lonelier than even a castaway lying on a small raft in the ocean. 
\ex
\glll
而 在 第二 天 拂晓 ， 当 一 个 奇怪 的 小 声音 叫醒 我 的 时候 ， 你们 可以 想见 我 当时 是 多么 吃惊 。
\\
er2 zai4 di1er4 tian1 fu2xiao3 , zai4 yi2 ge4 qi2guai4 de5 xiao3 sheng1yin1 jiao4xing2 wo3 de5 shi2hou3 , ni2men5 ke2yi3 xiang3jian4 wo3 dang1shi2 shi4 duo1ma5 chi1jing1 。
\\
but at second day morning , when one \textsc{clf} strange DE small voice {wake up} \textsc{1sg} de time , {\textsc{2pl}} can imagine \textsc{1sg} {at that time} was how surprised . 
\\
\trans But at the morning of the second day, when a strange, small voice woke me up, you can imagine how surprised I was at that time. 
\ex
\glll
这 小小 的 声音 说道 ：
\\
zhe4 xiao2xiao3 de5 sheng1yin1 shuo1dao3 ：
\\
this small DE voice say :
\\
\trans This small voice says:
\ex
\glll
“ 请 你 给 我 画 一 只 羊 ， 好吗 ？ ”
\\
“ qing2 ni3 gei2 wo3 hua4 yi4 zhi1 yang2 ， hao3ma1 ？ ”
\\
`` please \textsc{2sg} for \textsc{1sg} draw one \textsc{clf} sheep , yes ? ''
\\
\trans ``Please draw a sheep for me, okay?''
\ex
\glll
“ 啊 ！ ”
\\
“ a1 ！ ”
\\
`` ah ! ''
\\
\trans ``Ah!''
\ex
\glll
“ 给 我 画 一 只 羊 …… ”
\\
“ gei2 wo3 hua4 yi4 zhi1 yang2 …… ”
\\
`` for \textsc{1sg} draw one \textsc{clf} sheep ... ''
\\
\trans ``Draw a sheep for me...''
\ex
\glll
我 象是 受到 惊雷 轰击 一般 ， 一下子 就 站立 起来 。
\\
wo3 xiang4shi4 shou4dao4 jing1lei2 hong1ji1 yi1ban1 ， yi2xia4zi5 jiu4 zhan4li4 qi3lai2 。
\\
\textsc{1sg} like receive thunder hit {} , immediately just stand up .
\\
\trans As if I received a thunder hit, I stood up immediately. 
\ex
\glll
我 使劲 地 揉了揉 眼睛 ， 仔细 地 看了看 。
\\
wo2 shi3jin4 de5 rou2le5rou2 yan3jing1 ， zi3xi4 de5 kan4le5kan4 。
\\
\textsc{1sg} {with force} DE rubbed eye , careful DE looked .
\\
\trans I rubbed my eyes with force, and I looked carefully. 
\ex
\glll
我 看见 一 个 十分 奇怪 的 小 家伙 严肃 地 朝 我 凝眸 望 着 。
\\
wo2 kan4jian4 yi2 ge4 shi2fen1 qi2guai4 de5 xiao3 jia1huo5 yan2su4 de5 chao2 wo3 ning2mou2 wang4 zhe5 。
\\
\textsc{1sg} see one \textsc{clf} very strange DE small person serious DE towards me stare look ZHE .
\\
\trans I saw a very strange small person staring at me seriously.
\ex
\glll
这 是 后来 我 给 他 画 出来 的 最 好 的 一 副 画像 。
\\
zhe4 shi4 hou4lai2 wo2 gei3 ta1 hua4 chu1lai2 de5 zui5 hao3 de5 yi2 fu4 hua4xiang4 。
\\
this is afterwards \textsc{1sg} for \textsc{3sg.m} draw out DE most good DE one \textsc{clf} painting .
\\
\trans This is the best painting of him that I drew afterwards. 
\ex
\glll
可是 ， 我 的 画 当然 要 比 他 本人 的 模样 逊色 得 多 。
\\
ke3shi4 ， wo3 de5 hua4 dang1ran2 yao4 bi3 ta1 ben3ren2 de5 mo2yang4 xun4se4 de5 duo1 。
\\
but , \textsc{1sg} DE painting {of course} is {compared to} he himself DE appearance bad DE much .
\\
\trans But my painting is, of course, much worse than his appearance.
\ex
\glll
这 不 是 我 的 过错 。
\\
zhe4 bu2 shi4 wo3 de5 guo4cuo4 。
\\
this not is \textsc{1sg} DE fault .
\\
\trans this is not my fault.
\ex
\glll
六 岁 时 ， 大人 们 使 我 对 我 的 画家 生涯 失去 了 勇气 ， 除了 画 过 开 着 肚皮 和 闭 着 肚皮 的 蟒蛇 ， 后来 再 没有 学 过 画 。
\\
liu4 sui4 shi2 , da2ren2 men5 shi2 wo3 dui4 wo3 de5 hua4jia1 sheng1ya2 shi1qu4 le5 yong3qi4 , chu2le5 hua4 guo4 kai1 zhe5 du4pi2 he2 bi4 zhe5 du4pi2 de mang3she2 , hou4lai2 zai4 mei2you3 xue2 guo4 hua4 . 
\\
six {years old} time , adult \textsc{pl} make me regarding \textsc{1sg} DE painter career lose LE courage , except draw GUO open ZHE belly and close ZHE belly DE boa , afterwards again not learn GUO painting .
\\
\trans When I was six years old, the adults made me lose courage in my painter career, and, except for having drawn the boas with open and closed bellies, I did not learn to draw again.
\ex
\glll
我 惊奇 地 睁大 着 眼睛 看 着 这 突然 出现 的 小 家伙 。
\\
wo3 jing1qi2 de5 zheng1da4 zhe5 yan3jing1 kan4 zhe5 zhe4 tu1ran2 chu1xian4 de5 xiao3 jia1huo5 。
\\
\textsc{1sg} surprised DE {open widely} ZHE eyes see ZHE this suddenly appear DE small person .
\\
\trans I opened my eyes widely in surprise and looked at this small person who suddenly appeared. 
\ex
\glll
你们 不要 忘记 ， 我 当时 处 在 远离 人烟 千 里 之外 的 地方 。
\\
ni3men2 bu2yao4 wang4ji4 ， wo3 dang1shi2 chu3 zai4 yuan3li2 ren2yan1 qian1 li3 zhi1wai4 de5 di4fang1 。
\\
\textsc{2pl} {do not} forget , \textsc{1sg} {at that time} {located} at {away from} {human fire} thousand mile {outside} DE place .
\\
\trans Don't forget, I was at a place thousands of miles away from human habitation. 
\ex
\glll
而 这 个 小 家伙 给 我 的 印象 是 ， 他 既 不 象 迷 了 路 的 样子 ， 也 没有 半点 疲乏 、 饥渴 、 惧怕 的 神情 。
\\
er2 zhe4 ge4 xiao3 jia1huo3 gei2 wo3 de5 yin4xiang4 shi4 ， ta1 ji4 bu2 xiang4 mi2 le5 lu4 de5 yang4zi5 ， ye5 mei2you5 ban4dian3 pi2fa2 、 ji1ke3 、 ju4pa4 de5 shen2qing2 。
\\
but this \textsc{clf} small person give me DE impression is , \textsc{3sg.m} neither not {seem like} lost LE way DE manner , also {not have} {any bit} fatigue , {hunger and thirst} , fear DE {facial expression} .
\\
\trans But the impression that this small person gives me is that he neither looks like he is lost, nor does he have any look of fatigue, hunger, thirst, or fear. 
\ex
\glll
他 丝毫 不 象是 一 个 迷失 在 旷无人烟 的 大 沙漠 中 的 孩子 。
\\
ta1 si1hao2 bu2 xiang4shi4 yi2 ge4 mi2shi1 zai4 kuang4wu2ren2yan1 de5 da4 sha1mo4 zhong1 de5 hai2zi5 。
\\
\textsc{3sg.m} {at all} not {seem like} one \textsc{clf} lost at {empty with no human life} DE big desert inside de kid .
\\
\trans He doesn't look at all like a kid lost in a big desert away from human life. 
\ex
\glll
当 我 在 惊讶 之 中 终于 又 能 说 出 话 来 的 时候 ， 对 他 说道 ：
\\
dang1 wo3 zai4 jing1ya4 zhi1 zhong1 zhong1yu2 you4 neng2 shuo1 chu1 hua4 lai2 de5 shi2hou4 ， dui4 ta1 shuo1dao4 ：
\\
when \textsc{1sg} at surprise DE inside finally again can speak out words {to here} DE time , to \textsc{3sg.m} say :
\\
\trans When I can finally speak again in the middle of my surprise, I said to him :
\ex
\glll
“ 唉 ， 你 在 这儿 干 什么 ？ ”
\\
“ ai1 ， ni3 zai4 zhe4r gan4 shen3me5 ？ ”
\\
`` ah , \textsc{2sg} at here do what ? ''
\\
\trans ``Ah, what are you doing here?''
\ex
\glll
可是 他 却 不慌不忙 地 好象 有 一 件 重要 的 事 一般 ， 对 我 重复 地 说道 ：
\\
ke3shi4 ta1 que4 bu4huang1bu4mang2 de5 hao3xiang4 you3 yi2 jian4 zhong4yao4 de5 shi4 yi4ban1 ， dui4 wo3 chong2fu4 de5 shuo1dao4 ：
\\
but \textsc{3sg} yet {not hurried} DE {as if} have one \textsc{clf} important DE thing {} , to \textsc{1sg} repeated DE say :
\\
\trans But as if it is something important, he repeated to me unhurriedly :
\ex
\glll
“ 请 …… 给 我 画 一 只 羊 …… ”
\\
“ qing2 …… gei2 wo3 hua4 yi4 zhi1 yang2 …… ”
\\
`` please ... for \textsc{1sg} draw one \textsc{clf} sheep ... ''
\\
\trans ``Please... draw a sheep for me...''
\ex
\glll
当 一 种 神秘 的 东西 把 你 镇住 的 时候 ， 你 是 不 敢 不 听从 它 的 支配 的 ， 在 这 旷无人烟 的 沙漠 上 ， 面临 死亡 的 危险 的 情况 下 ， 尽管 这样 的 举动 使 我 感到 十分 荒诞 ， 我 还是 掏出 了 一 张 纸 和 一 支 钢笔 。
\\
dang1 yi4 zhong3 shen2mi4 de5 dong1xi5 ba2 ni3 zhen4zhu4 de5 shi2hou4 ， ni3 shi4 bu4 gan3 bu4 ting1cong2 ta1 de5 zhi1pei4 de5 ， zai4 zhe4 kuang4wu2ren2yan1 de5 sha1mo4 shang4 ， mian4lin2 si3wang2 de5 wei1xian3 de5 qing2kuang4 xia4 ， jin2guan3 zhe4yang4 de5 ju3dong4 shi2 wo3 gan3dao4 shi2fen1 huang1dan4 ， wo3 hai2shi4 tao1chu1 le5 yi4 zhang1 zhi3 he2 yi4 zhi1 gang1bi3 。
\\
when one \textsc{clf} mysterious DE thing BA you overpower DE time , you are not dare not {follow} \textsc{3sg.inanim} DE command DE , at this {empty with no human life} DE desert above , face death DE danger DE situation under , although such DE behavior make \textsc{1sg} feel very absurd , \textsc{1sg} still {take out} LE one \textsc{clf} paper and one \textsc{clf} pen .
\\
\trans When a mysterious thing overpowers you, you do not dare not to follow its command. At this empty desert with no human life, in a situation where one faces the danger of death, although such behavior made me feel very absurd, I still pulled out a piece of paper and a pen. 
\ex
\glll
这时 我 却 又 记起 ， 我 只 学 过 地理 、 历史 、 算术 和 语法 ， 就 有点 不大 高兴 地 对 小 家伙 说 我 不 会 画画 。
\\
zhe4shi2 wo3 que4 you4 ji4qi3 ， wo2 zhi3 xue2 guo4 di4li2 、 li4shi2 、 suan4shu4 he2 yu2fa3 ， jiu4 you2dian3 bu2da4 gao1xing4 de5 dui4 xiao3 jia1huo5 shuo1 wo3 bu2 hui4 hua4hua4 。
\\
{at this time} \textsc{1sg} but again remember , \textsc{1sg} only learn GUO geography , history , arithmetic and grammar , just {a bit} {not quite} happy DE to small person say \textsc{1sg} not can draw .
\\
\trans At this time I remembered, I only learned geography, history, arithmetic and grammar, so a bit not quite happily, I told the small person that I cannot draw.
\ex
\glll
他 回答 我 说 ：
\\
ta1 hui2da2 wo3 shuo1 ：
\\
\textsc{3sg} answer \textsc{1sg} say :
\\
\trans He answered me saying:
\ex
\glll
“ 没有 关系 ， 给 我 画 一 只 羊 吧 ！ ”
\\
“ mei2you3 guan1xi4 ， gei2 wo3 hua4 yi4 zhi1 yang2 ba5 ！ ”
\\
`` {no problem} {\_} , for \textsc{1sg} draw one \textsc{clf} sheep {} ! ""
\\
\trans ``No problem, draw a sheep for me!''
\ex
\glll
因为 我 从来 没有 画 过 羊 ， 我 就 给 他 重画 我 所仅仅 会 画 的 两 副 画 中 的 那 副 闭 着 肚皮 的 巨蟒 。
\\
yin1wei4 wo3 cong2lai2 mei2you3 hua4 guo4 yang2 ， wo2 jiu4 gei3 ta1 chong2hua4 wo2 suo3jin2jin2 hui4 hua4 de5 liang2 fu2 hua4 zhong1 de5 na4 fu2 bi4 zhe5 du4pi2 de5 ju4mang3 。
\\
because \textsc{1sg} ever not draw GUO sheep , \textsc{1sg} just for \textsc{3sg.m} redraw \textsc{1sg} only can draw DE two \textsc{clf} painting inside DE that \textsc{clf} close ZHE belly DE boa .
\\
\trans Because I have never drawn a sheep, I just redrew for him that boa with closed belly from the only two paintings that I can draw.
\ex
\glll
“ 不 ， 不 ！
\\
“ bu4 ， bu4 ！
\\
`` no , no !
\\
\trans ``No, no!
\ex
\glll
我 不 要 蟒蛇 ， 它 肚子 里 还 有 一 头 象 。 ”
\\
wo3 bu2 yao4 mang3she2 ， ta1 du4zi5 li3 hai2 you3 yi4 tou2 xiang4 。 ”
\\
\textsc{1sg} not want boa , \textsc{3sg.inanim} stomach inside also have one \textsc{clf} elephant . ''
\\
\trans I don't want a boa, who has an elephant inside its stomach.''
\ex
\glll
我 听 了 他 的 话 ， 简直 目瞪口呆 。
\\
wo3 ting1 le5 ta1 de5 hua4 ， jian3zhi2 mu4deng4kou3dai1 。
\\
\textsc{1sg} hear LE \textsc{3sg.m} DE words , simply stupefied .
\\
\trans I heard his words, and I was simply stupefied. 
\ex
\glll
他 接着 说 ： “ 巨蟒 这 东西 太 危险 ， 大象 又 太 占 地方 。
\\
ta1 jie1zhe5 shuo1  ： “ ju4mang3 zhe4 dong1xi5 tai4 wei1xian3 ， da4xiang4 you4 tai4 zhan4 di4fang1 。
\\
\textsc{3sg.m} continue say : `` boa this thing too dangerous , elephant also too occupy space . 
\\
\trans He continued: ``this boa thing is too dangerous, and the elephant takes up too much space.''
\ex
\glll
我 住 的 地方 非常 小 ， 我 需要 一 只 羊 。
\\
wo3 zhu4 de5 di4fang1 fei1chang2 xiao3 ， wo3 xu1yao4 yi4 zhi1 yang2 。
\\
\textsc{1sg} live DE place very small , \textsc{1sg} need one \textsc{clf} sheep .
\\
\trans The place I live is very small. I need a sheep.
\ex
\glll
给 我 画 一 只 羊 吧 。 ”
\\
gei2 wo3 hua4 yi4 zhi1 yang2 ba5 。 ”
\\
for \textsc{1sg} draw one \textsc{clf} sheep {} . ''
\\
\trans Draw a sheep for me.''
\ex
\glll
我 就 给 他 画 了 。
\\
wo3 jiu4 gei3 ta1 hua4 le5 。
\\
\textsc{1sg} just for \textsc{3sg.m} draw LE . 
\\
\trans So I drew for him.
\ex
\glll
他 专心 地 看 着 ， 随后 又 说 ：
\\
ta1 zhuan1xin1 de5 kan4 zhe5 ， sui2hou4 you4 shuo1 ：
\\
\textsc{3sg.m} careful DE look ZHE , afterwards also say :
\\
\trans He looked carefully, and then he said:
\ex
\glll
“ 我 不 要 ， 这 只 羊 已经 病 得 很 重 了 。
\\
“ wo3 bu2 yao4 ， zhe4 zhi1 yang2 yi3jing1 bing4 de5 hen4 zhong4 le5 。
\\
`` \textsc{1sg} not want , this \textsc{clf} sheep already sick DE very heavy LE . 
\\
\trans ``I don't want it. This sheep is already very sick. 
\ex
\glll
给 我 重新 画 一 只 。 ”
\\
gei2 wo3 chong2xin1 hua4 yi4 zhi1 。 ”
\\
for \textsc{1sg} again draw one \textsc{clf} . ''
\\
\trans draw one again for me.''
\ex
\glll
我 又 画 了 起来 。
\\
wo3 you4 hua4 le5 qi3lai2 。
\\
\textsc{1sg} again draw LE up .
\\
\trans I started drawing again.
\ex
\glll
我 的 这 位 朋友 天真可爱 地 笑 了 ， 并且 客气 地 拒绝道 ：
\\
wo3 de5 zhe4 wei4 peng2you3 tian1zhen1ke3ai4 de5 xiao4 le5 ， bing4qie3 ke4qi4 de4 ju4jue2dao4 ：
\\
\textsc{1sg} DE this \textsc{clf} friend {innocent and cute} DE laugh LE , and polite DE refuse :
\\
\trans This friend of mine laughed innocently, and refused politely:
\ex
\glll
“ 你 看 ， 你 画 的 不是 小羊 ， 是 头 公羊 ， 还 有 犄角 呢 。 ”
\\
“ ni3 kan4 ， ni3 hua4 de5 bu2shi4 xiao3yang2 ， shi4 tou2 gong1yang2 ， hai2 you3 ji1jiao3 ne5 。 ”
\\
`` \textsc{2sg} look , \textsc{2sg} draw DE {is not} sheep , is \textsc{clf} ram , also have horn {} . ''
\\
\trans ``Look, what you drew is not a sheep, it's a ram. It has horns.
\ex
\glll
于是 我 又 重新 画 了 一 张 。
\\
yu2shi4 wo3 you4 chong2xin1 hua4 le5 yi4 zhang1 。
\\
therefor \textsc{1sg} also again draw LE one \textsc{clf} .
\\
\trans Therefore I drew another one again.
\ex
\glll
这 副 画 同 前 几 副 一样 又 被 拒绝 了 。
\\
zhe4 fu2 hua4 tong2 qian2 ji3 fu2 yi2yang4 you4 bei4 ju4jue4 le5 。
\\
this \textsc{clf} painting as previous several \textsc{clf} same again BEI reject LE .
\\
\trans This painting, like the previous ones, was rejected. 
\ex
\glll
“ 这 一 只 太 老 了 。
\\
“ zhe4 yi4 zhi1 tai4 lao3 le5 。
\\
`` this one \textsc{clf} too old LE .
\\
\trans ``This one is too old.
\ex
\glll
我 想要 一 只 能 活 得 长 的 羊 。 ”
\\
wo3 xiang3yao4 yi4 zhi1 neng2 huo2 de5 chang2 de5 yang2 。 ”
\\
\textsc{1sg} want one \textsc{clf} can live DE long DE sheep . ''
\\
\trans I want a sheep that can live long. ''
\ex
\glll
我 不耐烦 了 。
\\
wo3 bu2nai4fan2 le5 。
\\
\textsc{1sg} {get impatient} LE .
\\
\trans I got impatient. 
\ex
\glll
因为 我 急于 要 检修 发动机 ， 于是 就 草草 画 了 这 张 画 ， 并且 匆匆 地 对 他 说道 ：
\\
yin1wei4 wo3 ji2yu2 yao4 jian3xiu1 fa1dong4ji1 ， yu2shi4 jiu4 cao2cao3 hua4 le5 zhe4 zhang1 hua4 ， bing4qie3 cong1cong1 de5 dui4 ta1 shuo1dao4 ：
\\
because \textsc{1sg} {be impatient} need repair motor , therefore just hastily draw LE this \textsc{clf} painting , and hurried DE to \textsc{3sg.m} say :
\\
\trans Because I am impatient to repair the motor, I just drew this painting hastily, and said to him hurriedly:
\ex
\glll
“ 这 是 一 只 箱子 ， 你 要 的 羊 就 在 里面 。 ”
\\
“ zhe4 shi4 yi4 zhi1 xiang1zi5 ， ni3 yao4 de5 yang2 jiu4 zai4 li3mian4 。 ”
\\
`` this is one \textsc{clf} box , \textsc{2sg} want DE sheep just {located at} inside . ''
\\
\trans ``This is a box. The sheep you want is inside.''
\ex
\glll
这时 我 十分 惊奇 地 看到 我 的 这 位 小 评判员 喜笑颜开 。
\\
zhe4shi2 wo3 shi2fen1 jing1qi2 de5 kan4dao4 wo3 de5 zhe4 wei4 xiao3 ping2pan4yuan2 xi3xiao4yan2kai1 。
\\
{at this time} \textsc{1sg} very surprise DE see \textsc{1sg} DE this \textsc{clf} small judge {smile happily} .
\\
\trans At this time, I was very surprised to see my small judge smiling happily.
\ex
\glll
他 说 ： “ 这 正 是 我 想要 的 ， ……
\\
ta1 shuo1 ： “ zhe4 zheng4 shi4 wo2 xiang3yao4 de5 ， ……
\\
\textsc{3sg.m} say : `` this exactly is \textsc{1sg} want DE , ...
\\
\trans He said: ``this is exactly what I wanted, ...
\ex
\glll
你 说 这 只 羊 需要 很多 草 吗 ？ ”
\\
ni3 shuo1 zhe4 zhi1 yang2 xu1yao4 hen3duo1 cao3 ma1 ？ ”
\\
you say this \textsc{clf} sheep need {a lot of} grass \textsc{ques} ? ''
\\
\trans Do you think that this sheep needs a lot of grass?''
\ex
\glll
“ 为什么 问 这个 呢 ？ ”
\\
“ wei4shen3me5 wen4 zhe4ge5 ne1 ？ ”
\\
`` why ask this {} ? ''
\\
\trans ``Why do you ask this?''
\ex
\glll
“ 因为 我 那里 地方 非常 小 …… ”
\\
“ yin1wei4 wo3 na4li3 di4fang1 fei1chang2 xiao3 …… ”
\\
`` because \textsc{1sg} there place very small ... ''
\\
\trans ``Because my place is very small...''
\ex
\glll
“ 我 给 你 画 的 是 一 只 很 小 的 小羊 ， 地方 小 也 够 喂养 它 的 。 ”
\\
“ wo2 gei2 ni3 hua4 de5 shi4 yi4 zhi1 hen2 xiao3 de5 xiao3yang2 ， di4fang1 xiao3 ye3 gou4 wei4yang2 ta1 de5 。 ”
\\
`` \textsc{1sg} for \textsc{2sg} draw DE is one \textsc{clf} very small DE sheep , place small also enough raise \textsc{3sg.inanim} DE . ''
\\
\trans ``What I drew for you is a very small sheep. Even a small place is enough to raise it.''
\ex
\glll
他 把 脑袋 靠近 这 张 画 。
\\
ta1 ba3 nao3dai4 kao4jin4 zhe4 zhang1 hua4 。
\\
\textsc{3sg.m} BA head approach this \textsc{clf} painting .
\\
\trans He moves his head close to this painting.
\ex
\glll
“ 并不 象 你 说 的 那么 小 ……
\\
“ bing4bu2 xiang4 ni3 shuo1 de5 na4me5 xiao3 ……
\\
`` not {similar to} \textsc{2sg} say DE that small ...
\\
\trans ``It's not as small as you said...
\ex
\glll
瞧 ！
\\
qiao2 ！
\\
look !
\\
\trans Look!
\ex
\glll
它 睡着 了 …… ”
\\
ta1 shui4zhao2 le5 …… ”
\\
\textsc{3sg.inanim} {fall asleep} LE ... ''
\\
\trans It fell asleep...''
\ex
\glll
就 这样 ， 我 认识 了 小 王子 。
\\
jiu4 zhe4yang4 ， wo3 ren4shi2 le5 xiao3 wang2zi5 。
\\
just {like that} , \textsc{1sg} {become acquainted with} LE small prince .
\\
\trans Just like that, I became acquainted with the Little Prince.
\end{exe}

\end{document}