%\documentclass[letterpaper,11pt]{article}
\documentclass[UTF8]{ctexart}
\usepackage[hang,flushmargin]{footmisc}

%\usepackage[empty]{fullpage}
\pagestyle{plain}

\setlength{\topmargin}{-15pt}
\setlength{\oddsidemargin}{0pt}
\setlength{\evensidemargin}{0pt}
\setlength{\textwidth}{460px}
\setlength{\voffset}{-1.5cm}
\setlength{\textheight}{23cm}
\setlength{\parindent}{0pt}
\setlength{\parskip}{10pt}

\nofiles

\usepackage{gb4e}

\setlength{\glossglue}{5pt plus 2pt minus 1pt}
\renewcommand{\eachwordthree}{\rule[-10pt]{0pt}{0pt}}

\begin{document}
\begin{center}
\textbf{\Large{Le Petit Prince - Glosses}}\\
%\vspace{3pt}
%\large{Sean Zhang}\\
%xiaoronz@andrew.cmu.edu
\end{center}


\begin{exe}
\ex 
\glll
当 我 还 只 有 六 岁 的 时候 ， 在 一 本 描写 原始 森林 的 名 叫 《 真实 的 故事 》 的 书 中 ， 看到 了 一 副 精彩 的 插画 ， 画 的 是 一 条 蟒蛇 正在 吞食 一 只 大 野兽 。\\
d\={a}ng w\v{o} h\'{a}i zh\'{i} y\v{o}u li\`{u} su\`{i} de sh\'{i}h\`{o}u , za\`{i} y\`{i} b\v{e}n mi\'{a}oxi\v{e} yu\'{a}nsh\v{i} s\={e}nl\'{i}n de m\'{i}ng ji\`{a}o 《 zh\={e}nsh\'{i} de g\`{u}sh\`{i} 》 de sh\={u} zh\={o}ng , k\`{a}nd\`{a}o le y\'{i} f\`{u} j\={i}ngc\v{a}i de ch\={a}hu\`{a} , hu\`{a} de sh\`{i} y\`{i} ti\'{a}o m\v{a}ngsh\'{e} zh\`{e}ngz\`{a}i t\={u}nsh\'{i} y\`{i} zh\={i} d\`{a} y\v{e}sh\`{o}u .\\
When \textsc{1sg} yet only have 6 years-old \textsc{gen} time , at one \textsc{clf} depict primitive forest \textsc{gen} name called 《 real \textsc{gen} story 》 \textsc{gen} book in , see \textsc{prf} one \textsc{clf} magnificent \textsc{gen} illustration , painted \textsc{gen} is one \textsc{clf} boa \textsc{prog} swallowing one \textsc{clf} big beast . \\
\trans When I was only 6 years old, in a book depicting primitive forests called ``real stories'', I saw a magnificent illustration showing a boa swallowing a giant beast. 
\ex
\glll
页头 上 就 是 那 副 画 的 摹本 。 \\
y\`{e}t\'{o}u sh\`{a}ng ji\`{u} sh\`{i} n\`{a} f\'{u} hu\`{a} de m\'{o}b\v{e}n .\\
{page header} on exactly is that \textsc{clf} painting \textsc{gen} copy .\\
\trans A copy of that painting is at the header of this page. 
\ex
\glll
这 本 书 中 写 道 ： `` 这些 蟒蛇 把 它们 的 猎获物 不 加 咀嚼 地 囫囵 吞下 ， 尔后 就 不 能 再 动弹 了 ； 它们 就 在 长长 的 六 个 月 的 睡眠 中 消化 这些 食物 。 '' \\
zhe ben shu zhong xie dao : `` zhexie mangshe ba tamen de liehuowu bu jia jujue de hulun tunxia , erhou jiu bu neng zai dongtan le ; tamen jiu zai changchang de liu ge yue de shuimian zhong xiaohua zhexie shiwu 。 ''
\\
this \textsc{clf} book inside write \textsc{comp} : `` these boa ``ba'' they \textsc{gen} prey no add chew \textsc{adv} complete swallow , afterwards then not {able to} {continue to} move \textsc{prf} ; they just at long \textsc{gen} six \textsc{clf} month \textsc{gen} sleep inside digest these food . ''
\\
\trans In this book it says: ``these boas swallow their preys as a whole, without chewing, afterwards they are no longer able to move; they then digest these foods during a long hibernation of six months. 
\ex
\glll
当时 ， 我 对 丛林 中 的 奇遇 想 得 很 多 ， 于是 ， 我 也 用 彩色 铅笔 画 出 了 我 的 第一 副 图画 。 \\
dangshi , wo dui conglin zhong de qiyu xiang de hen duo , yushi , wo ye yong caise qianbi hua chu le wo de diyi fu tuhua 。
\\
{at that time} , I regarding jungle inside \textsc{gen} adventure think \textsc{comp} very much , hence , I also use color pencil draw out \textsc{prf} I \textsc{gen} first \textsc{clf} painting .
\\
\trans At that time, I thought a lot about the adventures inside jungles, so I also used color pencils and drew my first painting. 
\ex
\glll
我 的 第一 号 作品 。 \\
wo de diyi hao zuopin 。 \\
I \textsc{gen} first \textsc{clf} work .
\\
\trans My first work. 
\ex
\glll
它 是 这样 的 ：\\
ta shi zheyang de : \\
it is {like this} \textsc{gen} :
\\
\trans It is like this:
\ex
\glll
我 把 我 的 这 副 杰作 拿 给 大人 看 ， 我 问 他们 我 的 画 是不是 叫 他们 害怕 。\\
wo ba wo de zhe fu jiezuo na gei daren kan , wo wen tamen wo de hua shibushi jiao tamen haipa 。\\
I ``ba'' I \textsc{gen} this \textsc{clf} masterpiece take to adult see , I ask them I \textsc{gen} painting {is or not} make them afraid .
\\
\trans I showed this masterpiece to the adults, and I asked them whether or not my painting makes them afraid.
\ex
\glll
他们 回答 我 说 ： “ 一 顶 帽子 有 什么 可怕 的 ？ ”\\
tamen huada wo shuo : `` yi ding maozi you shen keba de ? ''\\
they answer me saying : `` one \textsc{clf} hat has what scary \textsc{gen} ? ''
\\
\trans They answered me saying: ``what's so scary about a hat?''
\ex
\glll
我 画 的 不 是 帽子 ， 是 一 条 巨蟒 在 消化 着 一 头 大象 。\\
wo hua de bu shi maozi , shi yi tiao jumang zai xiaohuo zhe yi tou daxiang 。
\\
I draw \textsc{gen} not is hat , is one \textsc{clf} {giant boa} \textsc{prog} digest \textsc{prog} one \textsc{clf} elephant . 
\\
\trans What I drew is not a hat, it's a giant boa digesting an elephant. 
\ex
\glll
于是 我 又 把 巨蟒 肚子 里 的 情况 画 了 出来 ， 以 便 让 大人 们 能够 看懂 。\\
yushi wo you ba jumang duzi li de qingkuang hua le chulai , yi bian rang daren men nenggou kandong . 
\\
so I also ``ba'' boa stomach inside \textsc{gen} situation draw \textsc{prf} out , {} {so that} let adult \textsc{pl} {are able to} {understand} . 
\\
\trans So I also drew out the situation inside the giant boa's stomach, so that the adults can understand. 
\ex
\glll
这些 大人 总是 需要 解释 。\\
zhexie daren zongshi xuyao jieshi . 
\\
these adults always need explanations .
\\
\trans These adults always need explanations.
\ex
\glll
我 的 第二 号 作品 是 这样 的 ：\\
wo do dier hao zuopin shi zheyang de :
\\
I \textsc{gen} second \textsc{clf} work is {like this} \textsc{gen}:
\\
\trans My second work is like this:
\ex
\glll
大人 们 劝 我 把 这些 画 着 开 着 肚皮 的 ， 或 闭 上 肚皮 的 蟒蛇 的 图画 放 在 一边 ， 还是 把 兴趣 放 在 地理 、 历史 、 算术 、 语法 上 。\\
daren men quan wo ba zhexie hua zhe kai zhe dupi de ， huo bi shang dupi de mangshe de tuhua fang zai yibian ， haishi ba xingqu fang zai dili 、 lishi 、 suanshu 、 yufa shang 。
\\
adult \textsc{pl} advise me BA these draw ZHE open ZHE belly DE , or close up belly DE boa DE painting put at side , rather BA interest put at geography , history , arithmetic , grammar on .
\\
\trans The adults advised me to set aside the paintings of boas with belly open or with belly closed, and instead focus on geography, history, arithmetic, and grammar. 
\ex
\glll
就 这样 ， 在 六 岁 的 那 年 ， 我 就 放弃 了 当 画家 这 一 美好 的 职业 。\\
jiu zheyang ， zai liu sui de na nian ， wo jiu fangqi le dang huajia zhe yi meihao de zhiye 。
\\
just {like this} , at six {years old} DE that year , I just {give up} LE becoming painter this one wonderful DE career .
\\
\trans Just like that, at the age of six, I gave up the wonderful career of becoming a painter.
\ex
\glll
我 的 第一 号 、 第二 号 作品 的 不成功 ， 使 我 泄 了 气 。\\
wo de diyi hao 、 dier hao zuopin de buchenggong ， shi wo xie le qi 。
\\
I DE first \textsc{clf} , second \textsc{pl} work DE failure , make me disheartened LE {} .
\\
\trans The failure of my first and second works made me disheartened.
\ex
\glll
这些 大人 们 ， 靠 他们 自己 什么 也 弄 不 懂 ， 还得 老是 不断 地 给 他们 作 解释 。\\
zhexie daren men ， kao tamen ziji shenme ye nong bu dong ， haishi laoshi buduan de gei tamen zuo jieshi 。\\
these adult \textsc{pl} , {depend on} they self what also {} not understand , still always {non-stop} DE give them make explanation .
\\
\trans These adults, they understand nothing on their own, and I always explain things for them. 
\ex
\glll
这 真 叫 孩子 们 腻味 。\\
zhe zhen jiao haizi men niwei 。\\
this really make kid \textsc{pl} {fed up}.
\\
\trans This really makes the kids feel fed up. 
\ex
\glll
后来 ， 我 只好 选择 了 另外 一 个 职业 ， 我 学会 了 开 飞机 ， 世界 各地 差不多 都 飞 到 过 。
\\
houlai ， wo zhihao xuanze le lingwai yi ge zhiye ， wo xuehui le kai feiji ， shijie gedi chabuduo dou fei dao guo 。
\\
afterwards , I {can only} choose LE another one \textsc{clf} career , I learn LE drive plane , world everywhere almost all fly reach GUO .
\\
\trans Afterwards, I had to choose another career. I learned to fly a plane, and I've been to almost every place in the world. 
\ex
\glll
的确 ， 地理学 帮 了 我 很 大 的 忙 。
\\
dique ， dilixue bang le wo hen da de mang 。
\\
indeed , geography help LE me very big DE favor .
\\
\trans Indeed, geography helped me a lot. 
\ex
\glll
我 一眼 就 能 分辨 出 中国 和 亚里桑那 。
\\
wo yiyan jiu neng fenbian chu zhongguo he yalisangna 。
\\
I {in one glance} just can recognize China and Arizona .
\\
\trans I can recognize China and Arizona in one glance.
\ex
\glll
要是 夜里 迷失 了 航向 ， 这 是 很 有用 的 。
\\
yaoshi yeli mishi le hangxiang ， zhe shi hen youyong de 。
\\
if {at night} lose LE direction , this is very useful DE .
\\
\trans If I lost the direction at night, it is very useful. 
\ex
\glll
这样 ， 在 我 的 生活 中 ， 我 跟 许多 严肃 的 人 有 过 很 多 的 接触 。
\\
zheyang  ， zai wo de shanghuo zhong ， wo gen xuduo yansu de ren you guo hen duo de jiechu 。
\\
{like this} , in I DE life inside , I with many serious DE person have GUO very many DE contact .
\\
\trans Like that, in my life, I have had a lot of contact with many serious people. 
\end{exe}



\end{document}