% !TeX program = lualatex

\documentclass{book}
\usepackage{babel}
\babelprovide[main, import]{thai}
\babelfont{rm}{FreeSerif}
\usepackage{gb4e}

\begin{document}
	\begin{exe}
		\ex
		\gll เมื่อ ตอน ฉัน อายุ ได้ ๖ ขวบ ฉัน ได้ เห็น รูปภาพ จับใจ รูป หนึ่ง ใน หนังสือ เกี่ยวกับ ป่าดงดิบ ชื่อ-ว่า "ประวัติ ชีวิต ธรรมชาติ" \\
		 when at.time 1\textsc{SG} have.age.of \textsc{PST} six year.\textsc{CLF} 1\textsc{SG} \textsc{PST} see picture impress.\textsc{ADJ} picture.\textsc{CLF} \textsc{INDF} in book about rainforest name-\textsc{QUOT} "story life nature"\\
		\glt  When I was 6 years old, I saw an impressive picture in a book about the rainforest called "True Stories from Nature"
	\end{exe}

	\begin{exe}
	\ex
		\gll รูป นั้น เป็น รูป งูเหลือม กำลัง กลืน สัตว์ป่า \\
		picture.\textsc{CLF} \textsc{DEM} COP picture boa.constrictor \textsc{PRS}.\textsc{PTCP} swallow  wild.animal\\
		\glt  That picture was a picture of a boa constrictor swalling a wild animal.
	\end{exe}

	\begin{exe}
		\ex
		\gll นี่ คือ รูปลอก ของ ภาพ นั้น \\
		\textsc{DEM} \textsc{COP} sticker \textsc{POSS} image \textsc{DEM} \\
		\glt  This is a sticker(?) of that image.
	\end{exe}

	\begin{exe}
		\ex
		\gll ใน หนังสือ เขา บรรยาย ไว้ ว่า "งูเหลือม กลืน เหยื่อ ของ มัน เข้าไป ทั้งหมด โดย ไม่ เคี้ยว เลย\\
		in book \textsc{3SG} explain keep(?) \textsc{QUOT} boa swallow prey of it enter whole by not chew at.all\\
		\glt  In the book it explained, "The boa swallows its prey whole without chewing.
	\end{exe}

	\begin{exe}
		\ex
		\gll ถัด จากนั้น มัน ก็ ไม่ สามารถ เคลื่อนไหว ได้ และ นอน ย่อย อาหาร อยู่ ตลอด เวลา ๖ เดือน"\\
		next after.that \textsc{3SG} then \textsc{NEG} able.to move able.to and sleep digest food \textsc{PROG} throughout time six month\\
		\glt  After that, it is unable to move and sleeps while digesting for six months." 
	\end{exe}

	\begin{exe}
		\ex
		\gll ฉัน ติดใจ เรื่อง การผจญภัย ใน ป่าทึบ มาก\\
		\textsc{1SG} enthralled.by story adventure in thick.jungle very\\
		\glt  I was very enthralled by the story of adventure in the thick jungle.
	\end{exe}

	\begin{exe}
		\ex
		\gll จึง ได้ หัด วาดภาพ ด้วย ดินสอ สี จน สำเร็จ\\
		thus \textsc{PST} train draw \textsc{INS} pencil colored until succeed\\
		\glt  I then practiced drawing with a colored pencil until I succeeded.
	\end{exe}

	\begin{exe}
		\ex
		\gll นี่ คือ รูปภาพ รูป แรก ของ ฉัน\\
		this \textsc{COP} picture \textsc{CLF} first of \textsc{1SG}\\
		\glt  This is my first picture.
	\end{exe}

	\begin{exe}
		\ex
		\gll ฉัน อวด ผลงาน ชิ้น เยี่ยม นี้ แก่ ผู้ใหญ่ และ ถาม เขา ว่า รูป ของ ฉัน ทำให้ เขา กลัว ไหม\\
		\textsc{1SG} show.off product \textsc{CLF} excellent this to	 adult and ask \textsc{3SG} \textsc{QUOT} picture of \textsc{1SG} cause.to.be \textsc{3SG} fear \textsc{Q}\\
		\glt  I showed off my excellent work to the adults and asked them if it scared them.
	\end{exe}

	\begin{exe}
		\ex
		\gll พวกเขา ก็ ตอบ ว่า " ทำไม หมวก ทำให้ คน กลัว เล่า "\\
		\textsc{3PL} then answer \textsc{QUOT}  " why hat cause person fear (emphasis).\textsc{PART} "\\
		\glt  They answered, "Why on earth would a hat cause someone to fear?"
	\end{exe}

	\begin{exe}
		\ex
		\gll รูปภาพ ของ ฉัน ไม่ ได้ เป็น รูป หมวก\\
		picture of \textsc{1SG} not \textsc{PST} \textsc{COP} picture hat\\
		\glt  My picture was not a picture of a hat.
	\end{exe}

	\begin{exe}
		\ex
		\gll แต่ เป็น รูป งูเหลือม ที่ กำลัง นอน ย่อย ช้าง ที่ มัน กลืน เข้าไป\\
		but \textsc{COP} picture boa that \textsc{PRS.PTCP} sleep digest elephant that \textsc{3SG} swallow enter\\
		\glt  Rather it was a picture of a boa digesting an elephant that it had swallowed whole.
	\end{exe}

	\begin{exe}
		\ex
		\gll ฉัน จึง ต้อง วาด รูปภาย ใน งูเหลือม เพื่อ ให้ พวก ผู้ใหญ่ เข้าใจ เพราะ พวก นี้ ต้องการ คำ อธิบาย เสมอ\\
		\textsc{1SG} thus need.to draw picture inside boa so.that cause \textsc{PL} adult understand because \textsc{PL} this need word explain always\\
		\glt  I thus had to draw a picture inside a boa so the adults would understand, because they always need explanations.
	\end{exe}

	\begin{exe}
		\ex
		\gll รูปวาด ที่ ๒ ของ ฉัน จึง เป็น ดังนี้\\
		drawing number two of \textsc{1SG} thus \textsc{COP} as.such\\
		\glt Thus, my second drawing looked as such.
	\end{exe}

	\begin{exe}
		\ex
		\gll พวก-ผู้ใหญ่ ได้ แนะนำ ว่า ฉัน ควร-จะ เลิก ยุ่ง กับ การ-วาดภาพ งูเหลือม ชนิด ที่ เห็น ด้าน นอก หรือ ด้าน ใน เสีย และ หัน-มา สนใจ เรียน วิชา ภูมิศาสตร์ ประวัติศาสตร์ คณิตศาสตร์ และ ไวยากรณ์ แทน\\
		\textsc{PL}-adult \textsc{PST} advise that \textsc{1SG} should-\textsc{FUT} quit bother with \textsc{NMLZ}-draw boa style which see viewpoint outside or viewpoint inside completely.\textsc{PERF} and turn-come.\textsc{ALL} interested\_in learn study\_of geography history mathematics and grammar instead\\
		\glt The adults advised that I should stop bothering with drawing boas, inside or outside, and instead turn my interest towards the study of geography, history, mathematics, and grammar.
	\end{exe}

	\begin{exe}
		\ex
		\gll ดังนั้น ฉัน จึง ได้ ละทิ้ง อาชีพ วาด ภาพ อัน สูงส่ง นี้ เสีย ตั้งแต่ อายุ ๖ ขวบ\\
		consequently \textsc{1SG} thus \textsc{PST} abandon career draw picture \textsc{CLF} noble.majestic this completely.\textsc{PERF} since age six years\\
		Consequently, I completely abandoned the noble career of drawing at the age of six.
	\end{exe}

	\begin{exe}
		\ex
		\gll เพราะว่า ฉัน รู้สึก หมดกำลังใจ ที่ ภาพวาด รูป แรก และ รูป ที่สอง ไม่ ได้ รับ ผลสำเร็จ\\
		because \textsc{1SG} felt disheartened that drawing \textsc{CLF} first and \textsc{CLF} second not \textsc{PST} receive success\\
		For I was disheartened that my first and second drawings were not successful.
	\end{exe}

	\begin{exe}
		\ex
		\gll พวก-ผู้ใหญ่ ไม่ เคย เข้าใจ อะไร เลย\\
		\textsc{PL}-adult not ever understand anything at\_all\\
		Adults never understand anything.
	\end{exe}

	\begin{exe}
		\ex
		\gll เป็น เรื่อง ที่ น่า-เหนื่อยหน่าย อย่าง ยิ่ง ที่ พวกเรา เด็ก-ๆ จะ ต้อง อธิบาย ให้ พวกเขา เข้าใจ อยู่ ตลอดเวลา\\
		\textsc{COP} issue that \textsc{ADJ}-tired way extreme that \textsc{1PL} child-\textsc{PL} \textsc{FUT} have\_to explain such\_that \textsc{3PL} understand \textsc{PROG} throughout time\\
		It is extremely tiring for us children to always have to explain such that they can understand.
	\end{exe}

	\begin{exe}
		\ex
		\gll ดังนั้น ฉัน จึง ได้ เลือก อาชีพ ใหม่\\
		so \textsc{1SG} so \textsc{PST} choose career new\\
		So I then chose a new career.
	\end{exe}

	\begin{exe}
		\ex
		\gll ฉัน หัด ขับ เครื่อง บิน\\
		\textsc{1SG} train drive machine fly\\
		I trained to fly airplanes.
	\end{exe}

	\begin{exe}
		\ex
		\gll ฉัน บิน ไป เกือบ จะ ทั่ว โลก\\
		\textsc{1SG} fly go almost \textsc{FUT} whole world\\
		I've flown to almost the whole world.
	\end{exe}

	\begin{exe}
		\ex
		\gll และ ภูมิศาสตร์ ก็ ช่วย ฉัน มาก ทีเดียว\\
		and geography \textsc{PART}.(emphasis) help \textsc{1SG} a\_lot quite\\
		And geography has truly helped me quite a lot.
	\end{exe}

	\begin{exe}
		\ex
		\gll ฉัน สามารถ บอก ได้ ทันที ว่า นี่ คือ ประเทศ จีน ไม่ ใช่ แอริโซนา\\
		\textsc{1SG} able\_to tell able\_to immediately \textsc{QUOT} this \textsc{COP} country China not \textsc{COP} Arizona\\
		I can tell you immediately that this is China, not Arizona.
	\end{exe}

	\begin{exe}
		\ex
		\gll วิชา นี้ จึง เป็น ประโยชน์ มาก ถ้า เรา หลง ทาง ใน เวลา กลาง คืน\\
		knowledge this thus \textsc{COP} benefit a\_lot if \textsc{1SG} lose way in time middle\_of night\\
		This knowledge is a huge benefit if one loses their way in the middle of the night.
	\end{exe}

	\begin{exe}
		\ex
		\gll ตลอด ชีวิต ของ ฉัน ฉัน ต้อง ติดต่อ กับ คน เป็น จำนวน มาก\\
		Throughout life of \textsc{1SG} \textsc{1SG} have\_to communicate with person \textsc{COP} quantity many\\
		Throughout my life, I have had to communicate with a great number of people
	\end{exe}

	\begin{exe}
		\ex
		\gll เป็น คน ที่ เอาจริงเอาจัง ทั้งนั้น\\
		\textsc{COP} person who serious completely\\
		...People who are all very serious.
	\end{exe}

	\begin{exe}
		\ex
		\gll ฉัน อาศัย อยู่ ใน บ้าน กับ พวก ผู้ใหญ่\\
		\textsc{1SG} live located\_at in house with \textsc{PL} adult\\
		I have lived in a house with adults.
	\end{exe}

	\begin{exe}
		\ex
		\gll ฉัน ได้ ศึกษา คน พวก นี้ อย่าง ใกล้ ชิด\\
		\textsc{1SG} \textsc{PST} study person \textsc{PL} this way close near\\
		I have studied them intimately.
	\end{exe}

	\begin{exe}
		\ex
		\gll แต่ มัน็้ มิ ได้ ช่วย ให้ ฉัน มี ความเห็น ต่อ คน พวก นี้ ใน ด้าน ที่ ดี ขึ้น เลย\\
		But that(SIC) not(SIC) help such\_that \textsc{1SG} have opinion for person \textsc{PL} this in viewpoint which good more at\_all\\
		But that has not helped me have a better opinion of them.
	\end{exe}

	\begin{exe}
		\ex
		\gll เมื่อ ฉัน พบ ใครสักคน ที่ ฉัน เห็น ว่า พอ มี แวว จะ ฉลาด หลัก แหลม ฉัน ก็ ทด สอบ เขา ด้วย การให้ ดู รูปวาด รูป ที่ หนึ่ง ซึ่ง ฉัน ยังคง เก็บ รักษาไว้\\
		When \textsc{1SG} meet anyone who \textsc{1SG} see that enough have promise \textsc{FUT} be-clever  be-sharp \textsc{1SG} then quiz \textsc{3SG} also by giving look drawing \textsc{CLF} place one which \textsc{1SG} still keep save\\
		When I met anyone that looked to have the promise of being clever enough, I tested them by giving them my drawing number one which I still keep.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน อยาก ทราบ ว่า เขา เป็น คน ที่ เข้าใจ ใน สิ่ง ต่างๆ ได้ ดี หรือ ไม่\\
		\textsc{1SG} want know that \textsc{3DG} \textsc{COP} person who understand in things different able-to good or not\\
		I wanted to know if they were someone who could understand various things well or not.
	\end{exe}

	\begin{exe}
		\ex 
		\gll แต่ ทุก ครั้ง คน เหล่า นั้น จะ พูด ว่า " นี่ คือ หมวก "\\
		but every time person group this \textsc{FUT} say that \textsc{PUNCT} this \textsc{COP} hat \textsc{PUNCT}\\
		But every time those people would say, "this is a hat."
	\end{exe}

	\begin{exe}
		\ex 
		\gll เมื่อ เป็น อย่าง นั้น ฉัน จึง ไม่ ยอม พูด ถึง งูเหลือม ป่าดงดิบ หรือ ดวงดาว กับ เขา ให้ เสีย เวลา เลย\\
		When \textsc{COP} way than \textsc{1SG} thus not allow speak about boa rainforest or star with \textsc{3SG} such-that break time totally\\
		When that was the case, I would not waste time by speaking to them about boas, rainforests, or stars.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน จะ ปล่อย เขา ไป ตาม วิถีทาง ของ เขา\\
		\textsc{1SG} \textsc{FUT} release \textsc{3SG} go follow way of \textsc{3SG}\\
		I would let them follow their own way.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน จะ พูด กับ เขา เรื่อง ไพ่บริดจ์ เรื่อง กอล์ฟ เรื่อง การเมือง และ เรื่อง เนคไท\\
		\textsc{1SG} \textsc{FUT} speak with \textsc{3SG} about bridge about golf about politics and about neckties.\\
		I would speak with them about bridge, about golf, about politics and about neckties.
	\end{exe}

	\begin{exe}
		\ex 
		\gll และ พวก ผู้ใหญ่ เหล่า นั้น ก็ จะ ยินดี ที่ ได้ รู้จัก กับ คน ที่ พูด จา ได้ เรื่องราว คน หนึ่ง\\
		and \textsc{PL} adult group that then \textsc{FUT} pleased that \textsc{PST} meet with person who speak discuss able\_to matters \textsc{CLF} one\\
		And these adults would be pleased to meet someone who can speak on such matters.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ดังนั้น ฉัน จึง ดำรง ชีวิต อย่าง เดียว ดาย ไร้ คน ที่ จะ สนทนา กัน เข้า อก เข้า ใจ อย่าง แท้ จริง จน กระทั่ง เมื่อ ๖ ปี มา นี้ เครื่อง บิน ของ ฉัน เข้า ไป เครื่อง เสีย อยู่ กลาง ทะเล ทราย สะฮารา
		so \textsc{1SG} thus sustain life way alone lonely without person who \textsc{FUT} converse together enter chest enter heart way genuine true until until when six years come here machine fly of \textsc{1SG} enter go machine break \textsc{PROG} middle\_of sea sand Sahara\\
		So I lived alone, without anyone to talk to and truly empathize with me, until 6 years ago when my plane broke in the middle of the Sahara desert.\\
	\end{exe}

	\begin{exe}
		\ex 
		\gll มี อะไร บาง อย่าง ใน เครื่อง ยนต์ ขัดข้อง เนื่อง จาก ฉัน ไม่ มี ช่าง เครื่อง มา ด้วย ไม่ มี แม้ แต่ ผู้ โดยสาร ฉัน จึง จำ ต้อง ซ่อม ด้วย ตนเอง อยู่ คน เดียว อย่าง ยาก ลำบาก
		\textsc{EXIST} something some way in machine motor dissent due from \textsc{1SG} not have mechanic machine come also not have even but person be\_passenger \textsc{1SG} thus bind have\_to repair with self alone be person alone way difficult troublesome\\
		Something was wrong in the motor, and since I had no mechanic or even passengers, I was obliged to perform the difficult repairs by myself.\\
	\end{exe}

	\begin{exe}
		\ex 
		\gll นี่ เป็น ปัญหา ที่ ต้อง เอา ชีวิต ของ ฉัน เข้า เดิมพัน ที เดียว\\
		this \textsc{COP} problem which have\_to take life of \textsc{1SG} enter wager occasion single\\
		This was truly a problem on which I had to stake my life.
	\end{exe}

	\begin{exe}
		\ex 
		\gll เพราะ ฉัน มี น้ำ ไว้ ดื่ม อีก เพียง ๘ วัน เท่านั้น\\
		Because \textsc{1SG} have water keep drink additional only eight days only\\
		For I had only enough water to drink for eight more days.
	\end{exe}

	\begin{exe}
		\ex 
		\gll คืน แรก ฉัน นอน หลับ บน พื้น ทราย ห่าง ไกล จาก ผู้ คน นับ ล้าน ๆ ไมล์\\
		night first \textsc{1SG} lie\_down sleep on\_top\_of surface sand distant far from people people count million \textsc{EMPH} miles\\
		The first night, I slept on the sand millions of miles from civilization.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน อยู่ คน เดียว ประหนึ่ง เรือ แตก รอด อยู่ บน แพ เคว้ง คว้าง กลาง มหา สมุทร\\
		\textsc{1SG} stay person single as\_if boat smash survive stay on\_top\_of raft float drift middle\_of great ocean\\
		I was alone, as if I had survived a shipwreck and was adrift on a raft in the middle of the ocean.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ดังนั้น คุณ คง จะ นึก ออก ว่า ฉัน ตกใจ เพียง ใด ที่ ตอน ใน รุ่ง สาง ก็ มี เสียง เล็ก ๆ ปลุก ฉัน ขึ้น\\
		So \textsc{2SG} should \textsc{FUT} think exit that \textsc{1SG} shocked only any that time in dawn clear then \textsc{EXIST} sound small (tiny) wake \textsc{1SG} up\\
		So you can imagine how shocked I was at dawn to be woken by a tiny noise.
	\end{exe}

	\begin{exe}
		\ex 
		\gll เสียง นั้น กล่าว ว่า\\
		sound that said that\\
		That noise said
	\end{exe}

	\begin{exe}
		\ex 
		\gll " กรุณา วาด รูป แกะ ให้ ตัว หนึ่ง เถอะ ! "\\
		" please draw picture sheep \textsc{BEN} body.\textsc{CLF} one insist.\textsc{EMPH} "\\
		"Please draw me a sheep!"
	\end{exe}

	\begin{exe}
		\ex 
		\gll " หือ " \\
		" Huh "\\
		"Huh?"
	\end{exe}

	\begin{exe}
		\ex 
		\gll " ช่วย วาด รูป แกะ ให้ ฉัน ตัว หนึ่ง ซิ … "\\
		" help draw picture sheep \textsc{BEN} \textsc{1SG} body.\textsc{CLF} one please ... "\\
		"Please help draw me a sheep..."
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน ผวา ลุก ขึ้น ยืน ราว กับ ถูก สายฟ้า ฟาด\\
		\textsc{1SG} rush rise up stand similar as \textsc{PASS} lightning strike \\
		I jumped to my feet as if struck by lightning.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ยก มือ ขึ้น ขยี้ ตา มอง ดู อีก ที\\
		raise hand up rub eye look see another time\\
		I raised my hand to rub my eyes and look again.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ตา ฉัน ไม่ ได้ ฝาด ไป แน่\\
		eye \textsc{1SG} not able\_to deceive go certainly\\
		My eyes could certainly not have been deceiving me.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน เห็น เด็ก เล็ก ๆ คน หนึ่ง ลักษณะ แปลก มาก ยืน มอง ดู ฉัน อย่าง เคร่ง ขรึม\\
		\textsc{1SG} see child small \textsc{EMPH} person.\textsc{CLF} one appearance strange very stand look see \textsc{1SG} way severe solemn\\
		I saw a small child, who looked very strange, standing and looking at me with severity.
	\end{exe}

	\begin{exe}
		\ex 
		\gll นี่ คือ รูปภาพ ที่ ดี ที่สุด ที่ ฉัน จำลอง ขึ้น ภาย หลัง\\
		this \textsc{COP} picture that good most that \textsc{1SG} copy up side behind\\
		This is the best picture that I could copy later.
	\end{exe}

	\begin{exe}
		\ex 
		\gll แน่ ละ แม้ ว่า ภาพวาด ของ ฉัน ไม่ สวย สด เหมือน ตัว จริง ของ เขา\\
		certainly \textsc{PART} even_though that drawing of \textsc{1SG} not pretty fresh same body try of \textsc{3SG}\\
		Certainly, however, my drawing is not as handsome as his real person.
	\end{exe}

	\begin{exe}
		\ex 
		\gll แต่ มัน ก็ ไม่ใช่ ความผิด ของ ฉัน\\
		but \textsc{3SG}.\textsc{INANIMATE} not fault of \textsc{1SG}\\
		But that is not my fault.
	\end{exe}

	\begin{exe}
		\ex 
		\gll พวก ผู้ใหญ่ ได้ ทำ ให้ ฉัน หมดกำลังใจ ที่ จะ เอา ดี ทาง วาด รูป ตั้งแต่ อายุ ได้ ๖ ขวบ และ นอก จาก รูป งูเหลือม ชนิด เห็น ด้าน นอก และ ด้าน ใน แล้ว ฉัน ก็ ไม่ ได้ หัด วาด รูป อะไร อีก เลย\\
		\textsc{PL} adult \textsc{PST} make \textsc{BEN} \textsc{1SG} discourage that \textsc{FUT} take good way draw picture since have_age \textsc{PST} six year and outside from picture boa piece.\textsc{CLF} see side out and side in \textsc{PERF} \textsc{1SG} then not \textsc{PST} train draw picture any more at_all\\
		The adults discouraged me from pursuing drawing since I was 6 years old, and apart from the drawings of boas from the outside and the inside I have not trained drawing anything else at all.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน เบิ่ง ตา มอง ดู เด็ก น้อย นั้น อย่าง ประหลาดใจ\\
		\textsc{1SG} blank_stare eye look see child small that way dumbfounded\\
		I gazed dumbfoundedly at that small child.
	\end{exe}

	\begin{exe}
		\ex 
		\gll อย่า ลืม ว่า ฉัน อยู่ ไกล แสน ไกล จาก ดิน แดน ที่ มี ผู้คน อาศัย\\
		do_not forget that \textsc{1SG} located_at far extremely far from land region that have people reside\\
		Do not forget that I was extremely far any civilization.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ทว่า เด็ก น้อย นั้น ไม่ มี ทีท่า ว่า หลง ทาง เหนื่อย อ่อน หิวโหย กระหาย น้ำ หรือ เกรง กลัว แต่ อย่าง ใด\\
		however child small that not have manner that lose way fatigued soft hungry thirst water or fearful scared but way any (multiword: at all)\\
		But that small child didn't appear to be fatigued and lost, hungry, dying of thirst, or afraid at all.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ไม่ มี ท่า ทาง ว่า เป็น เด็ก หลง ทาง อยู่ กลาง ทะเล ทราย ห่าง ไกล ผู้คน นับ เป็น ร้อย เป็น พัน โยชน์ เลย\\
		Not have appearance way that \textsc{COP} child lose way located_at middle_of sea sand distant far people count \textsc{COP} hundred \textsc{COP} thousand mile\\
		He didn't appear to be a child lost in the desert many thousands of miles from civilization.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ใน ที่สุด ฉัน ก็ รวบ รวม สติ ได้ และ พูด ขึ้น ว่า\\
		in final \textsc{1SG} then collect gather sense able\_to and speak up that\\
		Finally, I was able to gather my senses and say,
	\end{exe}

	\begin{exe}
		\ex 
		\gll " แล้ว เธอ มา ทำ อะไร อยู่ ที่ นี่ หละ "\\
		" and \textsc{2SG} come do what \textsc{PROG} at here \textsc{EMPH} "\\
		"And what have you come here to do, hmm?"
	\end{exe}

	\begin{exe}
		\ex 
		\gll เขา ก็ พูด ซ้ำ อีก อย่าง อ่อน โยน เหมือน เป็น เรื่อง สำคัญ ว่า\\
		\textsc{3SG} then speak repeat another way soft tender same \textsc{COP} story important that\\
		He then repeated himself softly as if it were a matter of great importance, 
	\end{exe}

	\begin{exe}
		\ex 
		\gll " กรุณา วาด แกะ ให้ ฉัน ตัว หนึ่ง เถอะ "\\
		" please draw sheep \textsc{BEN} \textsc{1SG} body.\textsc{CLF} one \textsc{EMPH} "\\
		"Please draw me a sheep."
	\end{exe}

	\begin{exe}
		\ex 
		\gll ใน เมื่อ เรา รู้สึก ว่า มี สิ่ง ลี้ลับ เกิน กว่า ที่ จะ กล่าว ได้ เรา ก็ ไม่ กล้า ที่ จะ ขัด ขืน ต่อ มัน\\
		in when \textsc{1SG} feel that have thing mysterious exceeding more\_than which \textsc{FUT} utter able\_to \textsc{1SG} then not dare that \textsc{FUT} defy resist against \textsc{3SG}\\
		When one feels that they have something more mysterious than they could describe, one doesn't dare defy it.
	\end{exe}

	\begin{exe}
		\ex 
		\gll แม้ ว่า ฉัน จะ อยู่ ห่าง ไกล ผู้คน นับ ด้วย ร้อย โยชน พัน โยชน์ และ อยู่ ระหว่าง อันตราย ถึง ชีวิต ฉัน ก็ ดึง เอา กระดาษ และ ดินสอ ออก มา\\
		even\_though that \textsc{1SG} \textsc{FUT} located\_at distant far people count with hundred miles thousand miles and located\_at middle\_of danger about life \textsc{1SG} then pull take paper and pencil exit come\\
		Even though I was thousands of miles from civilization and my life was in danger, I pulled out paper and a pencil.
	\end{exe}

	\begin{exe}
		\ex 
		\gll แต่ แล้ว ฉัน ก็ นึก ขึ้น มา ได้ ว่า ฉัน เคย เรียน แต่ วิชา ภูมิศาสตร์ ประวัติศาสตร์ คณิตศาสตร์ และ ไวยากรณ์ เท่านั้น ฉัน จึง กล่าว กับ เด็ก น้อย ( อย่าง อารมณ์ ไม่ค่อย ดี ) ว่า ฉัน วาด ไม่ เป็น หรอก นะ\\
		But already \textsc{1SG} then recall up come able\_to that \textsc{1SG} ever study only subject geography history mathematics and grammar only \textsc{1SG} thus explain with child small ( way mood barely good ) that \textsc{1SG} draw not able\_to at\_all light.\textsc{PART}\\
		But immediately I recalled that I had never studied anything but geography, history, mathematics, and grammar, and I thus explained to the small child (hardly in a good mood) that I could not draw at all.
	\end{exe}

	\begin{exe}
		\ex 
		\gll เขา ตอบ ฉัน ว่า\\
		\textsc{3SG} answer \textsc{1SG} that\\
		He answered me,
	\end{exe}

	\begin{exe}
		\ex 
		\gll " ไม่ เป็น ไร หรอก\\
		" not \textsc{COP} anything at\_all\\
		"(multiword) That's no problem at all.
	\end{exe}

	\begin{exe}
		\ex 
		\gll วาด แกะ ให้ ฉัน ตัว ก็ แล้ว กัน "\\
		draw sheep \textsc{BEN} \textsc{1SG} body.\textsc{CLF} also then together "\\
		Just draw me a sheep (multiword) however you like."
	\end{exe}

	\begin{exe}
		\ex 
		\gll เนื่อง จาก ฉัน ไม่ เคย วาด แกะ เลย ฉัน จึง วาด ภาพ หนึ่ง ใน สอง ภาพ ที่ ฉัน วาด เป็น ให้ เขา\\
		due from \textsc{1SG} not ever draw sheep at\_all \textsc{1SG} thus draw picture one in two picture which \textsc{1SG} draw able\_to \textsc{BEN} \textsc{3SG}\\
		Since I had never drawn a sheep, I drew one of the two pictures that I knew how to draw for him.
	\end{exe}

	\begin{exe}
		\ex 
		\gll คือ ภาพ งูเหลือม ชนิด ที่ เห็น ด้าน นอก\\
		\textsc{COP} picture boa piece.\textsc{CLF} which see side out\\
		It was the picture of a boa from the outside.
	\end{exe}

	\begin{exe}
		\ex 
		\gll และ ฉัน ก็ ตก ตะลึง เมื่อ ได้ ยิน เขา ตอบ ว่า\\
		and \textsc{1SG} then fall stun when \textsc{PST} hear \textsc{3SG} answer that\\
		And I was stunned when I heard him answer,
	\end{exe}

	\begin{exe}
		\ex 
		\gll " ไม่ เอา\\
		" not want\\
		"I don't want that!
	\end{exe}

	\begin{exe}
		\ex 
		\gll ไม่ เอา\\
		not want\\
		I don't want that!
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน ไม่ อยาก ได้ ช้าง ใน งูเหลือม\\
		\textsc{1SG} not want receive elephant in boa\\
		I don't want to get an elephant inside of a boa.
	\end{exe}

	\begin{exe}
		\ex 
		\gll งูเหลือม น่ากลัว ออก และ ช้าง ก็ ตัว ใหญ่ เกะกะ เกิน ไป\\
		boa frightening exit and elephant also body large obstructive exceed go\\
		Boas are too frightening and elephants are too troublesomely large.
	\end{exe}

	\begin{exe}
		\ex 
		\gll บ้าน ฉัน เล็ก นิด เดียว\\
		house \textsc{1SG} small tiny alone\\
		My house is very tiny.
	\end{exe}

	\begin{exe}
		\ex 
		\gll SENTENCE THAT THAI DOESNT HAVE\\
		\\
		
	\end{exe}

	\begin{exe}
		\ex 
		\gll วาด แกะ ให้ ฉัน ตัว หนึ่ง เถอะ "\\
		draw sheep \textsc{BEN} \textsc{1SG} body.\textsc{CLF} one \textsc{EMPH} "\\
		Draw me a sheep!"
	\end{exe}

	\begin{exe}
		\ex 
		\gll ดังนั้น ฉัน จึง วาด แกะ ให้\\
		so \textsc{1SG} thus draw sheep \textsc{BEN}\\
		So I drew him a sheep.
	\end{exe}

	\begin{exe}
		\ex 
		\gll เขา มอง ดู อย่าง สนใจ และ วิจารณ์ ว่า\\
		\textsc{3SG} look see way interested and critique that\\
		He looked interesedly and critiqued,
	\end{exe}

	\begin{exe}
		\ex 
		\gll " ไม่ใช่\\
		" no\\
		"No.
	\end{exe}

	\begin{exe}
		\ex 
		\gll แกะ ตัว นี้ ไม่ สบาย มาก\\
		sheep \textsc{CLF} this not healthy a\_lot\\
		This sheep is not well.
	\end{exe}

	\begin{exe}
		\ex 
		\gll วาด ใหม่ อีก ตัว ซิ "\\
		draw new another \textsc{CLF} \textsc{EMPH} "\\
		Draw another one."
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน วาด ใหม่\\
		\textsc{1SG} draw new\\
		I drew another.
	\end{exe}

	\begin{exe}
		\ex 
		\gll เพื่อน ของ ฉัน หัวเราะ เบา ๆ\\
		friend of \textsc{1SG} laugh quiet \textsc{ADV}\\
		My friend laughed quietly.
	\end{exe}

	\begin{exe}
		\ex 
		\gll พูด อย่าง อ่อน น้อม ว่า\\
		speak way soft bow that\\
		(NOT IN FRENCH) He said politely,
	\end{exe}

	\begin{exe}
		\ex 
		\gll " เธอ คง จะ เห็น นะ ว่า มัน ไม่ใช่ แกะ แต่ เป็น แพะ\\
		" \textsc{2SG} should \textsc{FUT} see \textsc{PART} that it \textsc{NEG}.\textsc{COP} sheep but \textsc{COP} ram\\
		"You should be able to see that this is not a sheep, but a ram.
	\end{exe}

	\begin{exe}
		\ex 
		\gll เพราะ ว่า มัน มี เขา "\\
		because that it have horn "\\
		Because it has horns."
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน วาด ใหม่ อีก\\
		\textsc{1SG} draw new another\\
		Again, I drew another.
	\end{exe}

	\begin{exe}
		\ex 
		\gll แต่ เขา ก็ ปฏิเสธ อีก เช่น กัน\\
		but \textsc{3SG} also refuse another similar together\\
		But again he similarly refused.
	\end{exe}

	\begin{exe}
		\ex 
		\gll " ตัว นี้ มัน แก่ เกิน ไป\\
		\textsc{CLF} this it old exceed go\\
		"This one is too old.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน ต้องการ แกะ ที่ จะ ยัง มี ชีวิต อยู่ ต่อ ไป อีก นาน "\\
		\textsc{1SG} need sheep which \textsc{FUT} still have life stay continue go more long\_time "\\
		I need a sheep that still has a lot of life left."
	\end{exe}

	\begin{exe}
		\ex 
		\gll เมื่อ เป็น เช่น นั้น ฉัน ก็ หมด ความอดทน ประกอบ กับ ต้อง รีบ ซ่อม เครื่อง ยนต์ ฉัน จึง เขี่ย ๆ ภาพ นี้\\
		when \textsc{COP} like that \textsc{1SG} then finish patience combine with have\_to hurry repair machine motor \textsc{1SG} thus scratch \textsc{EMPH} picture this\\
		That, combined with the fact that I had to hurry to repair my engine, made me lose my patience and I scribbled down this picture.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน ยื่น ให้ เขา และ ว่า\\
		\textsc{1SG} hand\_over \textsc{BEN} \textsc{3SG} and say\\
		I gave it to him and said,
	\end{exe}

	\begin{exe}
		\ex 
		\gll " เอ้า นี่ คือ กรง\\
		" \textsc{EXCL} this \textsc{COP} cage\\
		"Here! This is a cage.
	\end{exe}

	\begin{exe}
		\ex 
		\gll แกะ ที่ เธอ ต้องการ อยู่ ใน กรง นี้ "\\
		sheep which \textsc{2SG} need located\_at in cage this "\\
		The sheep you need is in this cage."
	\end{exe}

	\begin{exe}
		\ex 
		\gll แต่ ฉัน ประหลาดใจ มาก ที่ เห็น ใบหน้า ของ เขา สดใส ขึ้น ทันที\\
		but \textsc{1SG} astonished very that see face of \textsc{3SG} brighten up immediately\\
		But I was astonished to see his face immediately light up.
	\end{exe}

	\begin{exe}
		\ex 
		\gll " เหมาะ ทีเดียว นี่ แหละ ที่ ฉัน อยาก ได้\\
		" correct quite this \textsc{EMPH} that \textsc{1SG} want receive\\
		"Exactly - this is what I wanted!
	\end{exe}

	\begin{exe}
		\ex 
		\gll เธอ คิด ว่า แกะ ตัว นี้ กิน หญ้า จุ ไหม "\\
		\textsc{2SG} think that sheep \textsc{CLF} this eat grass much \textsc{Q} "\\
		Do you think this sheep eats much grass?"
	\end{exe}

	\begin{exe}
		\ex 
		\gll " ทำไม หละ "\\
		" why \textsc{PART} "\\
		"Why?"
	\end{exe}

	\begin{exe}
		\ex 
		\gll " เพราะ ว่า บ้าน ฉัน มัน แคบ นะ ซี"\\
		" because that house \textsc{1SG} it narrow \textsc{PART} \textsc{PART}"\\
		"Because, you see, my house is very narrow."
	\end{exe}

	\begin{exe}
		\ex 
		\gll " คง พอ หรอก\\
		" probably enough of\_course\\
		"It will probably be enough, of course.
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน ให้ แกะ เธอ ตัว เล็ก นิด เดียว "\\
		\textsc{1SG} give sheep \textsc{2SG} \textsc{CLF} small tiny alone "\\
		I gave you a very tiny sheep."
	\end{exe}

	\begin{exe}
		\ex 
		\gll เขา ยื่น ศรีษะ มา ใกล้ รูปวาด\\
		\textsc{3SG} extend head come close drawing\\
		He brought his head close to the drawing.
	\end{exe}

	\begin{exe}
		\ex 
		\gll " คง ไม่ เล็ก เกิน ไป นะ …\\
		" probably not small exceed go \textsc{PART} ...\\
		"Probably not too small, yeah...
	\end{exe}

	\begin{exe}
		\ex 
		\gll โอ …\\
		oh ...\\
		Oh...
	\end{exe}

	\begin{exe}
		\ex 
		\gll มัน กำลัง หลับ ด้วย "\\
		it \textsc{PROG} sleep too "\\
		It's gone to sleep."
	\end{exe}

	\begin{exe}
		\ex 
		\gll ฉัน ก็ ได้ ทำ ความรู้จัก กับ เจ้า ชาย น้อย ด้วย ประการ ฉะนี้\\
		\textsc{1SG} then \textsc{PST} make acquiantence with prince boy small with type thusly\\
		And that was how I met the Little Prince.
	\end{exe}
\end{document}